Numerical and analytical solution of nonlinear electro-kinetic phenomena

### 2012-09-08 notes/work/ijnam.tex
LaTeX

\documentclass[final]{elsarticle}

%\usepackage{setspace}\doublespacing

\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{units}   
\usepackage{float}   
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage[colorlinks]{hyperref}
\usepackage{framed}
\usepackage{url}

@@

\journal{Journal of Computational Physics}

\begin{document}

\begin{frontmatter}

\title{Numerical and asymptotic solution of nonlinear electrokinetic phenomena}
\author[cs]{Roman Zeyde}
\author[cs]{Irad Yavneh}
\author[math]{Ehud Yariv}

\address[cs]{Department of Computer Science, Technion -- Israel Institute of Technology, 
Haifa 32000, Israel}
\address[math]{Department of Mathematics, Technion -- Israel Institute of Technology, 
Haifa 32000, Israel}

\begin{abstract}
A numerical solver for the simulation of ion-exchanging particles migration in an
electrolyte solution is presented. The electrokinetic transport process is
described by a nonlinear partial differential equation system, 
as well as nonlinear boundary conditions.
The nonlinear system is discretized and solved iteratively using Newton's method.
An asymptotic analytical solution is derived for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, and good correspondence is achieved.
\end{abstract}

\begin{keyword}
CFD \sep electrokinetics \sep asymptotics
\end{keyword}

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.2\textwidth]
            {figs/ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (blue) surface charge,
        it attracts negative (red) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

@@

The goal of this work is to develop and implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.
One such phenomenon is nonlinear electrokinetic vortex flow (as shown in Figure \ref{fig:Granule}), 
which can useful for microfluidic mixing applications \cite{wang2004mix, ben2002vortex},

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.4\textwidth]
            {figs/vortex.eps}
        \caption{A micro-vortex electrokinetic flow around a 1 mm spherical ion-exchanger granule. }
        \label{fig:Granule}
    \end{center}
\end{figure}

@@

The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$):
\begin{eqnarray*}
\bj^*_\pm &=& 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{eqnarray*}
where $D^*$ is the ionic diffusivity.

The electric potential $\varphi$ is normalized by the thermal voltage:
\begin{eqnarray*}
\varphi^* &=& \frac{k_B^* T^*}{z e^*} = \frac{R^* T^*}{z F^*},
\end{eqnarray*}
where $R^* = k_B^* N_A$ and $F^* = e^* N_A$ is the Faraday number.

The dimensional Stokes equations are given by:
\begin{eqnarray*}
\bnabla \cdot \bv &=& 0, \\
-\bnabla p^* + \mu^* \bLaplacian \bv^* + \eps^* \Laplacian \varphi^* \bnabla \varphi^* &=& 0.
\end{eqnarray*}

@@

The nondimensional Nernst-Plank equation is then written as:
\begin{eqnarray*}
\bj_\pm &=& 
-\bnabla c_\pm + \alpha \bv c_\pm \mp c_\pm \bnabla \varphi,
\end{eqnarray*}
where $\alpha = {v^* a^*}/{D^*} = {\eps^* (\varphi^*)^2}/{\mu^* D^*}$ 
is the dimensionless Peclet number.

The dimensional Poisson equation can be written as:
\begin{eqnarray*}
(c^*_+ - c^*_-) z F^* &=& -\eps^* \Laplacian \varphi^*,
\end{eqnarray*}
and after nondimensionalization, is given by:
\begin{eqnarray*}
c_+ - c_- &=& -\frac{\eps^* \varphi^*}{z F^* c^* (a^*)^2} \Laplacian \varphi = 
-\delta^2 \Laplacian \varphi,
\\
\delta^2 &=& \frac{\eps^* \varphi^*}{c^* F^* z (a^*)^2} = 
\pars{\frac{\delta^*}{a^*}}^2.
\end{eqnarray*}

@@

\subsubsection{Nernst-Planck Equations}
The ionic fluxes are given by ion diffusion, electrostatic forces and ion advection by the fluid:
\begin{eqnarray*}
  \bj_\pm &=& -\bnabla c_\pm \mp c_\pm \bnabla \varphi + \alpha \bv c_\pm,
\end{eqnarray*}

Due to conservation of ions, the fluxes are divergence-free:
\begin{eqnarray*}
\bnabla \cdot \bj_\pm &=& 0.
\end{eqnarray*}
These equation may be re-written by using the following transformations:
\begin{eqnarray*}
  c &=& \frac{c_+ + c_-}{2},\\
  q &=& \frac{c_+ - c_-}{2}.
\end{eqnarray*}
Thus, salt and charge fluxes are defined by:
\begin{eqnarray*}
  \bj &=& \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c, \\
  \bi &=& \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q,
\end{eqnarray*}
and are divergence-free as well:
\begin{eqnarray*}
\bnabla \cdot \bj &=& 0, \\
\bnabla \cdot \bi &=& 0. 
\end{eqnarray*}

\subsubsection{Incompressible Stokes Flow with Electrostatic Force}
Mass conservation (due to constant fluid density) reads:
\begin{eqnarray*}
\bnabla \cdot \bv &=& 0.
\end{eqnarray*}
Momentum conservation (force balance) reads:
\begin{eqnarray*}
\Laplacian \bv - \bnabla p + \bnabla \varphi \Laplacian \varphi &=& 0.
\end{eqnarray*}
Finally, free charge density is governed by the Poisson equation:
\begin{eqnarray*}
\delta^2 \Laplacian \varphi &=& -2q.
\end{eqnarray*}

@@

An ion-exchanger fixed reference frame is used.
Far away from the particle, we have a uniform applied electrostatic field, uniform flow,
and ambient ionic concentrations:
\begin{eqnarray*}
-\bv &\rightarrow& \cU \ui, \\
-\bnabla \varphi &\rightarrow& \beta\ui, \\
c_\pm &\rightarrow& 1.
\end{eqnarray*}

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we assume high conductance of the particle (yielding a uniform electric potential),
zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$):
\begin{eqnarray*}
\varphi & = & \cV, \\
\bv & = & \bzero, \\
\bn \cdot \bj_- & = & 0, \\
c_+ &=& \gamma, \\
\end{eqnarray*}
where $\gamma$ is the ratio between the cationic concentration on the surface and the cationic
concentration in the fluid bulk.

@@

\subsubsection  {Bulk-scale equations}
The Nernst-Planck equations above for the fluid bulk (outside
the boundary layer), using bulk variables denoted by uppercase letters, read:
\begin{eqnarray*}
  Q & = & 0, \\
  C & = & C_+ = C_-, \\
\bJ &=& -\bnabla C + \alpha \bV C, \\
\bI &=& -C \bnabla \varPhi.
\end{eqnarray*}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations:

\begin{enumerate}
\item Salt conservation in macroscale equation:
\begin{eqnarray*} \label{eq:salt}
\bnabla \cdot \bJ = \Laplacian C - \alpha \bV \cdot \bnabla C = 0. 
\end{eqnarray*}

\item Charge conservation in macroscale equations:
\begin{eqnarray*} \label{eq:charge}
\bnabla \cdot \bI = \bnabla \cdot \pars{ C \bnabla \varPhi } = 0.
\end{eqnarray*}

@@

The macroscale Stokes flow equations have the same form as the microscale equations:
\begin{eqnarray*}  \label{eq:stokes}
\brc{rcc}{\bnabla \cdot \bV &=& 0, \\  
\Laplacian \bV - \bnabla P + \bnabla \varPhi \Laplacian \varPhi &=& \bzero.}
\end{eqnarray*}

\subsubsection{Bulk-scale effective boundary conditions}
By integrating the microscale equations across the Debye layer
(see the full derivation in \cite{yariv2010asymptotic}), 
the effective boundary conditions at the ion-exchanger surface are:
\begin{eqnarray*}
0 &=& \varPhi + \log C, \\
0 &=& \bnhat \cdot \bnabla \pars{\varPhi - \log C}, \\
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C,
\end{eqnarray*}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential (the voltage drop across the Debye layer).

\subsection{Steady state}
The stress tensor is composed of Newtonian and Maxwel stresses:
\begin{eqnarray*}
\tT &=& \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi) \tI, \\
\bnabla \cdot \tT &=& \bzero.
\end{eqnarray*}
The total force acting on the particle is computed by the following surface integral:
\begin{eqnarray*}
\bF &=& \oint_\mathcal{S} \tT \cdot \bnhat dA.
\end{eqnarray*}
In steady state (constant particle drift velocity $\cU$), 
the total force must be zero: 
\begin{eqnarray*}
\bF = \bzero.
\end{eqnarray*}

@@

The steady-state velocity is thus the solution to the nonlinear scalar equation
\begin{eqnarray*}
F(\beta, \cU) = 0,
\end{eqnarray*}
which we solver by standard secant method.

\subsection{Discretization}
\subsubsection{Spherical Coordinates}
Since the system is axisymmetric, the operators are written in 
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{eqnarray*}
\bnabla f &=& \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, \\
\bnabla \cdot \bF &=& \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{eqnarray*}
The scalar Laplacian can be written as:
\begin{eqnarray*}
\Laplacian f = \bnabla \cdot (\bnabla f)&=& 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{eqnarray*}
The vector gradient can be written as:
\begin{eqnarray*}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat.
\end{eqnarray*}
The vector Laplacian can be written as:
\begin{eqnarray*}
\bLaplacian \bF &=& 
\left(\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}\right)\brhat
+ \left(\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}\right) \bthetahat.
\end{eqnarray*}

@@

\subsubsection{Computational Grid}
In order to discretize the differential operators, 
a regular grid of size $n_r \times n_\theta$ is defined, with:
\begin{eqnarray*}
(r_i,\theta_j) &\in& [1, \infty) \times [0,\pi], \\ 
r_i &=& (1+\Delta_r)^i, \\
\theta_j &=& \Delta_\theta \cdot j,
\end{eqnarray*}
where a logarithmic grid spacing has been chosen for $r$ and a uniform grid for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$ should be much 
larger than $r_0$. Thus:
\begin{eqnarray*}
\Delta_r &=& \pars{R_{max}} ^ \frac{1}{n_r} - 1, \\
\Delta_\theta &=& \frac{\pi}{n_\theta}.
\end{eqnarray*}

This grid induces a disjoint subdivision of the domain 
$\Omega = [1, R_{max}] \times [0,\pi]$ into cells:
\begin{eqnarray*}
\bigcup_{ij}\Omega_{ij} &=& \Omega.
\end{eqnarray*}
A specific cell $\Omega_{ij}$ and its center $(\bar{r}_i, \bar{\theta}_j)$ are defined by:
\begin{eqnarray*}
\Omega_{ij} &=& [r_{i-1}, r_{i}] \times [\theta_{j-1}, \theta_{j}], \\
\bar{r}_i &=& \frac{r_{i-1} + r_{i}}{2}, \\
\bar{\theta}_j &=& \frac{\theta_{j-1} + \theta_{j}}{2}.
\end{eqnarray*}

Each variable is discretized with respect of its cell as follows:
\begin{itemize}
\item $\varPhi$, $C$ and $P$ are represented by their value at the center of each cell, 
using an all-centered grid:
\begin{eqnarray*}
\varPhi^h_{[i,j]} &=& \varPhi(\bar{r}_i, \bar{\theta}_j), \\
C^h_{[i,j]} &=& C(\bar{r}_i, \bar{\theta}_j), \\
P^h_{[i,j]} &=& P(\bar{r}_i, \bar{\theta}_j).
\end{eqnarray*}
\item $\bV$ is represented by its values at cell boundaries, using a staggered grid:
\begin{eqnarray*}
V_r^h{}_{[i,j]} &=& V_r(r_i, \bar{\theta}_j), \\
V_\theta^h{}_{[i,j]} &=& V_\theta(\bar{r}_i, {\theta}_j).
\end{eqnarray*}
\end{itemize}

@@

Define the following discrete central difference operators:
\begin{eqnarray*}
\cD_r(f^h){}_{\left[i+\half,j\right]} &=& \frac{f^h_{\left[i+1,j\right]} - f^h_{\left[i,j\right]}}
                       {r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}, \\
\cD_\theta(f^h){}_{\left[i,j+\half\right]} &=& \frac{f^h_{\left[i,j+1\right]} - f^h_{\left[i,j\right]}}
					   {\theta_{\left[i,j+1\right]} - \theta_{\left[i,j\right]}}.
\end{eqnarray*}

>>>>

\documentclass[10pt]{ijnam}
\hsize=5.5 true in
\textheight=8.4 true in
\topmargin 1in

%% information for the editor
\def\currentvolume{1}
\def\currentissue{1}
\def\currentyear{2004}
\def\month{October}
\pagespan{1}{18}
\copyrightinfo{2012}{} % copyright year

%% please put your definitions into mydef.sty. Examples can be found in this
%% file.
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{units}   
\usepackage{float}   
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage[colorlinks]{hyperref}
\usepackage{framed}
\usepackage{url}

@@

%\usepackage{mydef}

%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%

\title{Numerical and analytical solution of nonlinear electro-kinetic phenomena}

% author and address
\author{Roman Zeyde \and Irad Yavneh}
\address{Department of Computer Science, Technion -- Israel Institute of Technology, Haifa 32000, Israel}
\email{romanz@cs.technion.ac.il \and irad@cs.technion.ac.il}

% two authors have the same address
\author{Ory Schnitzer \and Ehud Yariv}
\address{Department of Mathematics, Technion -- Israel Institute of Technology, Haifa 32000, Israel}
\email{oschnitt@gmail.com \and udi@math.technion.ac.il}

% dedication
\dedicatory{This paper is dedicated to our authors}

% communicated
\commby{Hilbert}

% Received by the editors ?
\date{January 1, 2004 and, in revised form, March 22, 2004.}

% it is suggested to put it in Acknowledgments
\thanks{This research was supported by ???}

\subjclass[2000]{35R35, 49J40, 60G40}

\abstract{A numerical solver for the simulation of ion-exchanging particles migration in an
electrolyte solution is presented. The electrokinetic transport process is
described by a nonlinear partial differential equation system, 
as well as nonlinear boundary conditions.
The nonlinear system is discretized and solved iteratively using Newton's method.
An asymptotic analytical solution is derived for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, and good correspondence is achieved.}

\keywords{CFD, electrokinetics, asymptotics}

\maketitle

\section{Introduction} \label{sec:intro}

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.2\textwidth]
            {figs/ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (blue) surface charge,
        it attracts negative (red) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

@@

The goal of this work is to develop and implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.
One such phenomenon is nonlinear electrokinetic vortex flow, 
which can useful for microfluidic mixing applications \cite{wang2004mix, ben2002vortex},

@@

The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$):
\begin{eqnarray}
\bj^*_\pm &=& 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{eqnarray}
where $D^*$ is the ionic diffusivity.

The electric potential $\varphi$ is normalized by the thermal voltage:
\begin{eqnarray}
\varphi^* &=& \frac{k_B^* T^*}{z e^*} = \frac{R^* T^*}{z F^*},
\end{eqnarray}
where $R^* = k_B^* N_A$ and $F^* = e^* N_A$ is the Faraday number.

The dimensional Stokes equations are given by:
\begin{eqnarray}
\bnabla \cdot \bv &=& 0, \\
-\bnabla p^* + \mu^* \bLaplacian \bv^* + \eps^* \Laplacian \varphi^* \bnabla \varphi^* &=& 0.
\end{eqnarray}

@@

The nondimensional Nernst-Plank equation is then written as:
\begin{eqnarray}
\bj_\pm &=& 
-\bnabla c_\pm + \alpha \bv c_\pm \mp c_\pm \bnabla \varphi,
\end{eqnarray}
where $\alpha = {v^* a^*}/{D^*} = {\eps^* (\varphi^*)^2}/{\mu^* D^*}$ 
is the dimensionless Peclet number.

The dimensional Poisson equation can be written as:
\begin{eqnarray}
(c^*_+ - c^*_-) z F^* &=& -\eps^* \Laplacian \varphi^*,
\end{eqnarray}
and after nondimensionalization, is given by:
\begin{eqnarray}
c_+ - c_- &=& -\frac{\eps^* \varphi^*}{z F^* c^* (a^*)^2} \Laplacian \varphi = 
-\delta^2 \Laplacian \varphi,
\\
\delta^2 &=& \frac{\eps^* \varphi^*}{c^* F^* z (a^*)^2} = 
\pars{\frac{\delta^*}{a^*}}^2.
\end{eqnarray}

@@

\subsubsection{Nernst-Planck Equations}
The ionic fluxes are given by ion diffusion, electrostatic forces and ion advection by the fluid:
\begin{eqnarray}
  \bj_\pm &=& -\bnabla c_\pm \mp c_\pm \bnabla \varphi + \alpha \bv c_\pm,
\end{eqnarray}

Due to conservation of ions, the fluxes are divergence-free:
\begin{eqnarray}
\bnabla \cdot \bj_\pm &=& 0.
\end{eqnarray}
These equation may be re-written by using the following transformations:
\begin{eqnarray}
  c &=& \frac{c_+ + c_-}{2},\\
  q &=& \frac{c_+ - c_-}{2}.
\end{eqnarray}
Thus, salt and charge fluxes are defined by:
\begin{eqnarray}
  \bj &=& \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c, \\
  \bi &=& \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q,
\end{eqnarray}
and are divergence-free as well:
\begin{eqnarray}
\bnabla \cdot \bj &=& 0, \\
\bnabla \cdot \bi &=& 0. 
\end{eqnarray}

\subsubsection{Incompressible Stokes Flow with Electrostatic Force}
Mass conservation (due to constant fluid density) reads:
\begin{eqnarray}
\bnabla \cdot \bv &=& 0.
\end{eqnarray}
Momentum conservation (force balance) reads:
\begin{eqnarray}
\Laplacian \bv - \bnabla p + \bnabla \varphi \Laplacian \varphi &=& 0.
\end{eqnarray}
Finally, free charge density is governed by the Poisson equation:
\begin{eqnarray}
\delta^2 \Laplacian \varphi &=& -2q.
\end{eqnarray}

@@

An ion-exchanger fixed reference frame is used.
Far away from the particle, we have a uniform applied electrostatic field, uniform flow,
and ambient ionic concentrations:
\begin{eqnarray}
-\bv &\rightarrow& \cU \ui, \\
-\bnabla \varphi &\rightarrow& \beta\ui, \\
c_\pm &\rightarrow& 1.
\end{eqnarray}

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we assume high conductance of the particle (yielding a uniform electric potential),
zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$):
\begin{eqnarray}
\varphi & = & \cV, \\
\bv & = & \bzero, \\
\bn \cdot \bj_- & = & 0, \\
c_+ &=& \gamma, \\
\end{eqnarray}
where $\gamma$ is the ratio between the cationic concentration on the surface and the cationic
concentration in the fluid bulk.

@@

\subsubsection  {Bulk-scale equations}
The Nernst-Planck equations above for the fluid bulk (outside
the boundary layer), using bulk variables denoted by uppercase letters, read:
\begin{eqnarray}
  Q & = & 0, \\
  C & = & C_+ = C_-, \\
\bJ &=& -\bnabla C + \alpha \bV C, \\
\bI &=& -C \bnabla \varPhi.
\end{eqnarray}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations:

\begin{enumerate}
\item Salt conservation in macroscale equation:
\begin{eqnarray} \label{eq:salt}
\bnabla \cdot \bJ = \Laplacian C - \alpha \bV \cdot \bnabla C = 0. 
\end{eqnarray}

\item Charge conservation in macroscale equations:
\begin{eqnarray} \label{eq:charge}
\bnabla \cdot \bI = \bnabla \cdot \pars{ C \bnabla \varPhi } = 0.
\end{eqnarray}

@@

The macroscale Stokes flow equations have the same form as the microscale equations:
\begin{eqnarray}  \label{eq:stokes}
\brc{rcc}{\bnabla \cdot \bV &=& 0, \\  
\Laplacian \bV - \bnabla P + \bnabla \varPhi \Laplacian \varPhi &=& \bzero.}
\end{eqnarray}

\subsubsection{Bulk-scale effective boundary conditions}
By integrating the microscale equations across the Debye layer
(see the full derivation in \cite{yariv2010asymptotic}), 
the effective boundary conditions at the ion-exchanger surface are:
\begin{eqnarray}
0 &=& \varPhi + \log C, \\
0 &=& \bnhat \cdot \bnabla \pars{\varPhi - \log C}, \\
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C,
\end{eqnarray}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential (the voltage drop across the Debye layer).

\subsection{Steady state}
The stress tensor is composed of Newtonian and Maxwel stresses:
\begin{eqnarray}
\tT &=& \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi) \tI, \\
\bnabla \cdot \tT &=& \bzero.
\end{eqnarray}
The total force acting on the particle is computed by the following surface integral:
\begin{eqnarray}
\bF &=& \oint_\mathcal{S} \tT \cdot \bnhat dA.
\end{eqnarray}
In steady state (constant particle drift velocity $\cU$), 
the total force must be zero: 
\begin{eqnarray}
\bF = \bzero.
\end{eqnarray}

@@

The steady-state velocity is thus the solution to the nonlinear scalar equation
\begin{eqnarray}
F(\beta, \cU) = 0,
\end{eqnarray}
which we solver by standard secant method.

\subsection{Discretization}
\subsubsection{Spherical Coordinates}
Since the system is axisymmetric, the operators are written in 
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{eqnarray}
\bnabla f &=& \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, \\
\bnabla \cdot \bF &=& \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{eqnarray}
The scalar Laplacian can be written as:
\begin{eqnarray}
\Laplacian f = \bnabla \cdot (\bnabla f)&=& 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{eqnarray}
The vector gradient can be written as:
\begin{eqnarray}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat.
\end{eqnarray}
The vector Laplacian can be written as:
\begin{eqnarray}
\bLaplacian \bF &=& 
\left(\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}\right)\brhat
+ \left(\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}\right) \bthetahat.
\end{eqnarray}

@@

\subsubsection{Computational Grid}
In order to discretize the differential operators, 
a regular grid of size $n_r \times n_\theta$ is defined, with:
\begin{eqnarray}
(r_i,\theta_j) &\in& [1, \infty) \times [0,\pi], \\ 
r_i &=& (1+\Delta_r)^i, \\
\theta_j &=& \Delta_\theta \cdot j,
\end{eqnarray}
where a logarithmic grid spacing has been chosen for $r$ and a uniform grid for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$ should be much 
larger than $r_0$. Thus:
\begin{eqnarray}
\Delta_r &=& \pars{R_{max}} ^ \frac{1}{n_r} - 1, \\
\Delta_\theta &=& \frac{\pi}{n_\theta}.
\end{eqnarray}

This grid induces a disjoint subdivision of the domain 
$\Omega = [1, R_{max}] \times [0,\pi]$ into cells:
\begin{eqnarray}
\bigcup_{ij}\Omega_{ij} &=& \Omega.
\end{eqnarray}
A specific cell $\Omega_{ij}$ and its center $(\bar{r}_i, \bar{\theta}_j)$ are defined by:
\begin{eqnarray}
\Omega_{ij} &=& [r_{i-1}, r_{i}] \times [\theta_{j-1}, \theta_{j}], \\
\bar{r}_i &=& \frac{r_{i-1} + r_{i}}{2}, \\
\bar{\theta}_j &=& \frac{\theta_{j-1} + \theta_{j}}{2}.
\end{eqnarray}

Each variable is discretized with respect of its cell as follows:
\begin{itemize}
\item $\varPhi$, $C$ and $P$ are represented by their value at the center of each cell, 
using an all-centered grid:
\begin{eqnarray}
\varPhi^h_{[i,j]} &=& \varPhi(\bar{r}_i, \bar{\theta}_j), \\
C^h_{[i,j]} &=& C(\bar{r}_i, \bar{\theta}_j), \\
P^h_{[i,j]} &=& P(\bar{r}_i, \bar{\theta}_j).
\end{eqnarray}
\item $\bV$ is represented by its values at cell boundaries, using a staggered grid:
\begin{eqnarray}
V_r^h{}_{[i,j]} &=& V_r(r_i, \bar{\theta}_j), \\
V_\theta^h{}_{[i,j]} &=& V_\theta(\bar{r}_i, {\theta}_j).
\end{eqnarray}
\end{itemize}

@@

Define the following discrete central difference operators:
\begin{eqnarray}
\cD_r(f^h){}_{\left[i+\half,j\right]} &=& \frac{f^h_{\left[i+1,j\right]} - f^h_{\left[i,j\right]}}
                       {r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}, \\
\cD_\theta(f^h){}_{\left[i,j+\half\right]} &=& \frac{f^h_{\left[i,j+1\right]} - f^h_{\left[i,j\right]}}
					   {\theta_{\left[i,j+1\right]} - \theta_{\left[i,j\right]}}.
\end{eqnarray}

Define the following linear interpolation operators:
\begin{eqnarray}
\cI_r(f^h)_{\left[i+\half,j\right]} &=& 
\frac{r_{\left[i+\half,j\right]} - r_{\left[i,j\right]}}
{r_{\left[i+1,j\right]} - r_{\left[i,j\right]}} 
f^h_{\left[i+1,j\right]} 
+ 
\frac{r_{\left[i+1,j\right]} - r_{\left[i+\half,j\right]}}
{r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}
f^h_{\left[i,j\right]},
\\
\cI_\theta(f^h)_{\left[i,j+\half\right]} &=& 
\frac{r_{\left[i,j+\half\right]} - r_{\left[i,j\right]}}
{r_{\left[i,j+1\right]} - r_{\left[i,j\right]}}
  f^h_{\left[i,j+1\right]} + 
\frac{ r_{\left[i,j+1\right]} - r_{\left[i,j+\half\right]}}
{r_{\left[i,j+1\right]} - r_{\left[i,j\right]}}
  f^h_{\left[i,j\right]}.
\end{eqnarray}

The total flux of a cell is equal to zero ($\bnabla \cdot \boldsymbol{f} = 0$):
\begin{eqnarray}
\frac{1}{r^2} \cD_r\pars{f^h_r r^2} + 
\frac{1}{r \sin\theta} \cD_\theta\pars{f^h_\theta \sin\theta} =
\frac{\cD_r\pars{f^h_r \cdot r^2 \sin\theta} + 
\cD_\theta\pars{f^h_\theta \cdot r \sin\theta}}{r^2 \sin\theta} = 0. 
\end{eqnarray}
